---
title: "Project 1 Exploratory Analysis"
author: "Jack Cunningham"
format: pdf
editor: visual
---

# Loading packages and data

```{r}
library(tidyverse)
load("data_processed.RData")
```

# Response variables

Starting with numeric variables.

Starting first with eggs laid:

```{r}
data_processed |> 
  ggplot(aes(Num_Eggs_laid,result)) +
  geom_boxplot()
```

We can see that the number of eggs laid is a steady zero for when the tick dies. If the tick dies during oviposition the number of eggs laid is generally lower than when the tick lives. The variation in the number of eggs is also greater when the tick lives, compared to oviposition death. There exists one outlier value from the number of eggs being laid exceeds 6000. Lets look at that row:

```{r}
data_processed |> 
  filter(Num_Eggs_laid > 6000)
```

There does not appear to be anything suspicious about this observation, and it may indicate some right skew in the number of eggs laid variable. Generally, this is a symmetric distribution however.

Let's look at the eggs laid across the vaccination groups:

```{r}
data_processed |> 
  ggplot(aes(Num_Eggs_laid,Vaccine_Group)) +
  geom_boxplot()
```

Interestingly enough it appears that when the vaccine dose is low the lower number of eggs are laid. The control and the higher vaccine dose seem quite similar when it comes to this metric.

This motivates me to take a look at another response variable we have, number of larvae. Lets see how this varies across vaccine groups:

```{r}
data_processed |> 
  ggplot(aes(Num_Larvae,result)) +
  geom_boxplot()
```

It seems we have discovered a couple of values that seem particularly in feasible. Lets look at those rows:
