---
title: "Project 1 Exploratory Analysis"
author: "Jack Cunningham"
format: pdf
editor: visual
---

```{r, loading packages}
library(tidyverse)
```

# Loading Data

```{r}
data <- readxl::read_excel("Project2_S25_Data.xlsx")[,1:12]
```

```{r}
data
```

# Missing Data

```{r}
na_count <- sapply(data, function(y) sum(length(which(is.na(y)))))
na_count
```

We can see that there does not appear to be a lot of missing data on first glance, but after looking at the excel we can see that there does appear to be many "N/A" values. I also filter out the Comments column, because after reviewing the excel it seems it was only populated when there were extreme circumstances.

```{r}
data_no_coms <- data[,1:11]
na_count_manual <- sapply(data_no_coms, function(y) sum(length(which(y == "N/A"))))
na_count_manual
```

There appears to be a significant amount of data that is manually classified with "N/A". Lets see if they are often in the same row.

```{r}
data$na_count <- apply(data_no_coms, 1, function(x) sum(x == "N/A"|is.na(x)))
table(data$na_count)
```

We can see that almost always missing values are grouped in the same row.

# Deer groups:

Client said there are three deer groups.

```{r}
table(data$Deer_ID)
```

We see 12 different Deer IDs, mostly evenly distributed data points. Will need to figure out how these are grouped. These deer have different vaccination statuses:

```{r}
data |> group_by(Deer_ID, Vaccine_Group) |> 
  summarise(n = n()) |> 
  View()
```

In some instances we have deer with the same ID but different vaccination status. This makes me question what the Deer ID variable means.

# Vaccination Status

```{r}
table(data$Vaccine_Group)
```

There are four vaccine statuses. P - Pre-infested, C - Control, L - low, H - high.

Lets take another look at the Deer_ID/Vaccinate status combinations with that in mind:

```{r}
data |> group_by(Deer_ID, Vaccine_Group) |> 
  summarise(n = n()) |> 
  View()
```

We can see that the duplicates in Deer_ID/Vaccination Status are due to that pre-infested group.

# Infestation

Next we look at the Infestation feature.

```{r}
table(data$Infestation)
```

We can see that the count decreases as the stage of infestation grows further, from 478 to 304.
