---
title: "Calyx Data Exploratory Analysis"
format: html
editor: visual
---

```{r}
load("calyx_data.RData")
library(tidyverse)
```

# Seeing distribution of Measurement

```{r}
calyx_data |> 
  ggplot(aes(x = measurement)) + 
  geom_density()
```

The overall distribution of measurement has two peaks, which suggests there are two different distributions. Let's look at the length split into the start and end of the experiment.

```{r}
calyx_data |> 
  ggplot(aes(x = measurement)) +
  geom_density()+
  facet_wrap(~time)
```

At time 0 the measurement variable seems to be a symmetric distribution.

```{r}
calyx_data |> 
  filter(time == 0) |> 
  ggplot(aes(sample = measurement)) +
  stat_qq() +
  stat_qq_line()
```

This follows the normal distribution pretty closely.

```{r}
calyx_data |> 
  filter(time == 0) |>
  pull(measurement) |> 
  shapiro.test()
```

And the shapiro-wilk test agrees.

# Looking at T35

```{r}
calyx_data |> 
  filter(time == 0) |> 
  ggplot(aes(x = measurement, col = treatment)) + 
  geom_density()
```

```{r}
calyx_data |> 
  filter(time == 34) |> 
  ggplot(aes(x = measurement, col = treatment)) + 
  geom_density()
```

There seems to be evidence of the distribution of measurement being shifted depending on treatment. In particular the two with temperature of 20 are shifted to the right.

# Looking at T0 to T35 Split by Group

```{r}
calyx_data |>
  group_by(time,treatment) |> 
  summarise(avg_measurement = mean(measurement)) |> 
  ggplot(aes(x = time, y = avg_measurement)) +
  geom_col() +
  facet_wrap(~ treatment)
```
