---
title: "DNA Data Exploratory Analysis"
author: "Jack C"
format: html
editor: visual
---

Loading Data/Packages

```{r}
library(tidyverse)
load("DNA_data.RData")
```

# Looking at Cq at time 0 versus 35 for Polyps

```{r}
DNA_data |>
  filter(Sample_Type == "Polyp") |> 
  ggplot(aes(x = Cq, col = time)) +
  geom_density()
```

We can see that just looking at time leads to an almost uniform distribution for time 35, lets see how each treatment differs.

Before plotting, lets get an idea of the number of observations for each group:

```{r}
DNA_data |> 
  filter(time == 35 & Sample_Type == "Polyp") |> 
  group_by(treatment) |> 
  summarize(avg_Cq = mean(Cq), n = n())
```

```{r}
DNA_data |> 
  filter((time == 35) &(Sample_Type == "Polyp")) |> 
  ggplot(aes(x = Cq, col = treatment)) +
  geom_boxplot()
```

The differences between samples seem to be significantly different than each other for the polyp data.

# Looking at Cq at time 35 for Water

```{r}
DNA_data |> 
  filter(Sample_Type == "Water") |> 
  ggplot(aes(x = Cq, col = treatment)) +
  geom_boxplot()
```

There is moderate evidence that the treatment affects the DNA sample of water.
