---
title: "Calyx Data Confirmatory Analysis"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(car)
load("calyx_data.RData")
```

# Are all T0 samples the same

Getting t = 0 data:

```{r}
t0_calyx_data <- calyx_data |> 
  filter(time == 0)

```

To confirm that the samples do not differ based on the treatments they are assigned to we run an anova test:

```{r}
t0_model <- lm(measurement ~ treatment ,data = t0_calyx_data)
anova(t0_model)
```

We can see that the distribution is symmetric and appears close to normal:

```{r}
par(mfrow = c(1,2))
t0_calyx_data |> 
  ggplot(aes(x = measurement)) + 
  geom_density()

t0_calyx_data |> 
  ggplot(aes(sample = measurement)) +
  geom_qq() +
  geom_qq_line()
```

For a formal test of whether this data comes from a normal distribution we do a shapiro-wilk test:

```{r}
t0_calyx_data |> 
  pull(measurement) |> 
  shapiro.test()
```

Which does not reject the null hypothesis that this data was sampled from a normal distribution.

So there is no evidence for the T0 samples being drawn from different distributions.

# Are the T35 samples significantly different from one another across treatments?

Getting T35 data:

```{r}
t35_calyx_data <- calyx_data |> 
  filter(time == 34)
```

```{r}
t35_calyx_data |>    
  ggplot(aes(x = measurement, col = treatment)) +    
  geom_density()
```

There seems to be evidence of the distribution of measurement being shifted depending on treatment. In particular the two with temperature of 20 are shifted to the right.

We test this formally by creating a linear model and using anova to spot differences:

```{r}
t34_model <- lm(measurement ~ treatment ,data = t35_calyx_data)
anova(t34_model)
```

We can reject the null hypothesis of no difference due to the treatment applied the 5% confidence level.

# Are there differences between T0 and T35 between temperatures and microbe treatments?

```{r}
calyx_data |>
  group_by(time,treatment) |> 
  summarise(avg_measurement = mean(measurement)) |> 
  ggplot(aes(x = time, y = avg_measurement)) +
  geom_col() +
  facet_wrap(~ treatment)
```

It appears that differences are seen after treatments are applied, but not totally sure what analysis would be appropriate.

```{r}
t0_t35_diff_model <- lm(measurement ~ time*temperature*water, data = calyx_data)
Anova(t0_t35_diff_model)
summary(t0_t35_diff_model)
```

Would we need to show the three way interaction being significant?
