---
title: "Ephyra Data Confirmatory Analysis"
author: "Jack Cunningham"
format: html
editor: visual
---

# Loading Data and Packages:

```{r}
library(tidyverse)
load("ephyra_data.RData")
```

# Introduction

To begin, only four of the six treatments saw Ephyra transitions. For temperature 20 there were no Ephyra. So those two treatments will be excluded from the analysis of count and width.

# Are there differences in Width for Ephyra

```{r}
ephyra_data |> 
  filter(width != 0) |> 
  ggplot(aes(x = treatment, y = width)) + 
  geom_boxplot()
```

We exclude width measurements of zero as the missing values do not appear to have a pattern.

```{r}
ephyra_data_no_zero_width <-  ephyra_data |> 
  filter(width != 0)
```

A cursory view shows that it appears temperature has a large impact on the width of ephyra. To verify this we model:

```{r}
width_model <- lm(width ~ temperature*water, data = ephyra_data_no_zero_width)
anova(width_model)
```

We can see that temperature is significant in determining the eventual width of the Ephyra.

# Are there differences in Count of Ephyra by treatment

```{r}
ephyra_count_data <-  ephyra_data |> 
  filter(width != 0) |> 
  group_by(temperature, water, replicate) |> 
  summarise(n = n())
```

The count of ephyra seems to depend mainly on water. Lets see if our model confirms that:

```{r}
count_model <- glm(n ~ temperature*water, data = ephyra_count_data,
                   family = poisson())
anova(count_model)
summa
```
